Interfacing logic programming and Coq

### 2020-10-26 papers/types2020/intro.tex
Split out another two sections.

%% Important dates:
%%  Paper submission: 19 October 31 October 2020
%%  Author notification: 18 January 2021
%%  Final version:  15 February 2021
%%  Publication (presumably): end of March 2021

%% Style requirements of LIPIcs.  The recommended length of a paper is
%% 12-15 pages, excluding front-page(s) (authors, affiliation, keywords,
%% abstract, ...), bibliography and an appendix of max 5 pages. If you
%% need more pages, please ask the editors.

%% Papers have to be submitted in pdf through EasyChair:
%% https://easychair.org/conferences/?conf=types2020postproceed

\documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
\input head
\input preamble

\title{Interfacing logic programming and Coq} % Other suggestions?
\titlerunning{Interfacing logic programming and Coq}

\begin{document}
\maketitle

\begin{abstract}
We do stuff
\end{abstract}

@@

Recently, C. Sacerdoti Coen and E. Tassi developed the ELPI
implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol}.
More recently, Tassi has made ELPI available as a plugin to the Coq
proof assistant \cite{tassi18coqpl}.  This implementation of \lP
extends earlier implementations in primarily two directions.  First,
ELPI adds a notion of constraints, which makes it more expressive than,
say, the Teyjus implementation of \lP \cite{nadathur99cade}.  Second,
the plugin version of ELPI comes equip with an API for accessing the
Coq environment and a quotation and anti-quotation syntax allowing for
Coq expressions to be mixed in with \lP program elements.

@@

\section{Two cultures}

In proof theory, one learns quickly that things always come in pairs:
negative/positive, left/right, bottom-up/top-down,
premises/conclusion, introduction/elimination,
% determinism/nondeterminism,
% invertible/noninvertible, 
etc.
%
When we examine the larger setting of this project of linking a logic
programming engine with Coq and its kernel, we find a large number of
pairing that are valuable to explicitly describe.

\subsection{Proof theory vs type theory}

In many ways, (structural) proof theory is more elementary and
low-level than most approaches to type theory.  For example, type
theories usually answer the question ``What is a proof?'' with the
response ``a dependently typed $\lambda$-term''.  That is, when
describing a type theory, one usually decides that a proof is a
certain kind of term within the system.  In contrast, the proof theory
treats logical propositions and proofs as separate.  For example,
proof theory does not assume that there are terms within the logic
that describe proofs.

\begin{metanote}
  This paragraph seem to identify proof and type theory with sequents and nat ded and while
  I agree with the gist (although people such as Frank would argue that sequents are a way of organizing proof search for nat ded), we're not gonna make a lot of friends at types and more importantly is a big hammer to nail the point that logic programming is helpful with Coq --am 
\end{metanote}

Gentzen's discovery that the key to treating classical and
intuitionistic logics in the same proof system was the invention of
the weakening and contraction structural rules on the right-side of
sequents~\cite{gentzen35}.  This discovery lead him away from natural
deduction to the invention of multiple-conclusion sequent calculus, a
very important kind of proof that has not been given a satisfactory
representation in type theory.  This same innovation of Gentzen also
opened the way to the invention of another key proof theoretic
discovery, that of linear logic \cite{girard87tcs}.
Similarly, sequent calculus provides an elegant presentation for
linear logic while most treatments of linear logic in type theory
usually only capture linearity in a partial fashion.

Proof theory also has rather immediate and natural treatments for
co-induction (as well as induction: see
\cite{baelde12tocl,momigliano12jal}) whereas the treatment of
coinduction within type theory remains more challenging and is well
developed only for various syntactic conditions (see, for example,
\cite{bertot08entcs}).
\begin{metanote}
  The criticism applies to Coq, less to type theory, see co-patterns --am
\end{metanote}

\subsection{Proof search vs proof normalization}

Gentzen-style proofs are used to model computation in at least two
different ways.  The functional programming paradigm, following the
Curry-Howard correspondence, models computation abstractly as the
$\beta$-reduction of natural deduction proof: that is, computation is
modeled using \emph{proof normalization}.  On the other hand, the
logic programming paradigm, following the notion of goal-directed
proof search, models computation abstractly as a regimented search of
cut-free sequent calculus proofs: that is, computation is modeled
using \emph{proof search}.

Proof search has proved to have some richness that is hard to capture
in the proof normalization setting.  In particular, Gentzen's
invention of \emph{eigenvariables} are a kind of a proof-level
binder.  In the proof normalization setting, such eigenvariables
are instantiated during proof normalization.  In contrast, during the
search for cut-free proofs, eigenvariables are part of the syntax of
terms and formulas.  As a result, they can be used in the treatment of
bindings in data structures more generally.  Hence, the logic
programming paradigm provides a completely natural, elegant, and
powerful method, called $\lambda$-tree syntax, for treating bindings
within syntax.

In this paper, we shall use two different proof systems.  The first
logic is first-order intuitionistic logic.  Here, predicates
are not inductively defined and quantification will be limited to
range over first-order term.  Actually, we shall examine only the Horn
clause fragment of that logic.
%
The second logic will allow for inductively defined predicates using a
fixed point notation.   Given again our restriction to essentially
Horn clause specifications, it is possible to view intuitionistic
reasoning with such specification as if they are actually
specification in $\mu$MALL, an extension of multiplicative additive
linear logic with the least and greatest fixed point operators
\cite{baelde12tocl,baelde07lpar,heath19jar}.
%
The difference between these two logics and associate proof theories
is reminiscent to the difference between the type \lsti{Prop}
(corresponding to undefined propositions) and Sets (corresponding to
inductively defined propositions).
\begin{metanote}
  Still not onboard with this --am
\end{metanote}

\subsection{\lP vs Coq}

Given that \lP and Coq are both examples of mixing the
$\lambda$-calculus with logic, it is important to understand their
differences.  The confusion around the term \emph{higher-order
abstract syntax} (HOAS), for example, helps to illustrate their
differences.  In the functional programming setting, including the Coq
system, the HOAS approach leads to encoding binding structures within
terms using functions.  Such encodings have results in surprisingly
complicated encodings, allowing, for example, for \emph{exotic terms}
\cite{despeyroux95tlca} and to structures on which induction is not
possible \cite{roeckl01fossacs}.  In the logic programming setting,
particularly \lP, HOAS is well supported in both the availability of
binders allowed with terms ($\lambda$-bindings), in formulas
(quantifiers), and in proofs (eigenvariables).  For this reason, the
term $\lambda$-tree syntax was introduced to name this particular take
on HOAS \cite{miller19jar}.  The Abella proof assistant
\cite{baelde14jfr} was designed, in part, to allow for inductive and
co-inductive proofs for specifications using the $\lambda$-tree syntax
approach.

Another difference between \lP and functional programming can be
illustrated by considering how they are used in the specification of
tactics that are popular in modern proof assistants.  In fact, the
origin story for the ML functional programming language was that ML
was the meta-language for implementing the LCF suite of tactics and
tacticals \cite{gordon79}.  In order to implement tactics, ML adopted
not only novel features such as polymorphically typed higher-order
functional programming but also the non-functional mechanisms of
failure and exception handling.  While \lP is also based on
polymorphically typed higher-order relational programming, it also
comes with a completely declarative version of failure and
backtracking search.  Combining those features along with its support
of unification (even in the presence of term-level bindings), \lP
provides a rather different approach to the specification of tactics
\cite{felty93jar}.

While \lP is a typed language, it only uses simple types and these are
more often used to indicate syntactic categories.  Any dependency
information needs to be captured using predicates and quantifiers.

\input sec3

\input pbt

\input elab

\section{Conclusion}

\bibliography{l}
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

% LocalWords:  Tassi ELPI Teyjus intuitionistic Gentzen's Gentzen LCF
% LocalWords:  eigenvariables tacticals polymorphically Sacerdoti
% LocalWords:  Coen sequents

>>>>

@@

Recently, C. Sacerdoti Coen and E. Tassi developed the ELPI
implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol}.
More recently, Tassi has made ELPI available as a plugin to the Coq
proof assistant \cite{tassi18coqpl}.  This implementation of \lP
extends earlier implementations in primarily two directions.  First,
ELPI adds a notion of constraints, which makes it more expressive than,
say, the Teyjus implementation of \lP \cite{nadathur99cade}.  Second,
the plugin version of ELPI comes equip with an API for accessing the
Coq environment and a quotation and anti-quotation syntax allowing for
Coq expressions to be mixed with \lP program elements.

@@

### 2020-10-26 papers/types2020/main.tex
some small progress on sec 4
Split out another two sections.

\section{Introduction}
\label{sec:intro}

Recently, C. Sacerdoti Coen and E. Tassi developed the ELPI
implementation \cite{dunchev15lpar} of \lP 
\cite{miller12proghol}.  More recently, Tassi has made ELPI available
as a plugin to the Coq proof assistant \cite{tassi18coqpl}.
%
This implementation of \lP extends earlier implementations in
primarily two directions.  First, ELPI adds a notion of contraints,
which makes it more expressive than, say, the Teyjus implementation of
\lP \cite{nadathur99cade}.  Second, the plugin version of ELPI comes
equip with an API for accessing the Coq environment and a quotation
and anti-quotation syntax allowing for Coq expressions to be mixed in
with \lP program elements.

ELPI provides a means for treating bindings and unification (of
structures containing bindings) that are convenient for various
meta-programming tasks in logic programming all things useful for
tactic programming in Coq (proof refinement).  For examples,
ELPI programmers are spared from dealing with low-level aspects of
representing bindings (e.g., De Bruijn indexes) while still having
clean and effective ways to manipulate binding structures.
%
In short, ELPI appears to be a useful meta-language for Coq
\cite{coen19mscs,tassi18coqpl,tassi19itp}.
%
In this paper, we outline two applications of the Coq plugin for ELPI
that support this claim.
%
With these examples, we shall illustrate that ELPI is a useful
meta-language not just because for its treatment of bindings and
syntactic structures in general but also because it is actually based
on a sound implementation of a higher-order intuitionistic logic.
%
Type inference is a well-known example of a meta-programming task for
which logic programming often provides immediate and elegant
implementations via the sound implementation of logic.
%
In this paper, we present two additional project in which an
implementation of proof search provides direct, elegant, and compact
implementations of meta-programming tasks.
%
Before we present these two examples, we first discuss the different
world views that the integration of ELPI and Coq focuses us to
confront.

\section{Two cultures}

{\color{red} The rest of this section are just notes.}

These applications are not necessarily focused on the underlying
declarative nature of lambda Prolog.  For example, ELPI has various
extensions, such as ``spilling'' that are designed to make some
aspects of logic programming appear to be more functional.

In this paper, we limit our attention mostly to using lambda Prolog
for declarative processing: i.e., where logic plays a central role in
describing operations.  (Type checking is an example of such
processing, but we do not focus on type checking here).

Of course, being able to do declarative and not-so-declarative
operations within the same language makes that language all the more
useful.

  Given that lP and Coq are both examples of mixing the
  lambda-calculus with logic, it is important to understand their
  difference, something that can be very confusing.  Take the
  confusion around the term HOAS.  In some circles, this is a
  appropriate and sensible approach.  In other circles, it is
  semantically questionable.  But, in fact, this concept (using
  meta-level binding for object-level binding) plays our 
  differently depending on the meta language!

  The origin story to ML (Meta language) is that it supported
  LFC-style tactics and tacticals.  The strong typing of ML provided
  some guarantees: only theorems were ever given type thm.  But many
  steps of evolution have now taken place since then.  Isabelle's
  different approach to tactics.  Logic and proof theory, when used
  with Church's simple theory of types and Gentzen's sequent calculus,
  provides a far richer meta language: that provides an immediate and
  declarative treatment of bindings as well as for proof search
  (without the need of exception handling, etc).  [I'm making the case
  that FP is not a particularly natural framework for proof
  assistants.  Of course, lots of libraries have now been built in the
  FP style.]

  - Various dichotomies we find in this work.
    - proof theory vs type theory
    - logic programming vs functional programming
    - sequent calculus vs natural deduction

  - While both lambda Prolog and Coq are based on an amalgamation of
    $\lambda$-calculus and logic, these two languages take strikingly
    different approaches to bindings.  In lambda Prolog, bindings are
    syntactic and extensionality at any type is not available.  In
    Coq, however, higher-order typed quantification encodes
    extensional functions.  In lambda Prolog, (exists w. [lambda x. x]
    = [lambda x. w]) fails, no matter the type of w.  If this query
    was about functions, then this equation would be true when the
    type of w is a singleton.  But in lP, this is a query only about
    binding structures, and this fails to be proved since there is no
    capture-avoiding substitution (the only substitution supported by
    Church's STT) for w that makes the resulting two terms alpha,
    beta, eta convertible.  [Is this true in Coq?  What is equality
    here?]

  - On the proof theory side of this work, we use two different proof
    systems.

    - LJF, here we have undefined predicates, no induction
    - muMALL, were we have inductive definitions of predicates

    The Coq world provides for both of these kinds of predicate using
    Prop (corresponds to undefined) and Set (inductively defined).

\section{Elaboration of external proof certificates for the Coq
  kernel}

Follow the text in \cite{blanco20coq}.

\section{Property based testing for Coq}

@@

% Not part of git repo.  Need to find a fix
\bibliography{../../../../repo/parsifal/papers/references/master}

\end{document}

>>>>

\input intro
\input cultures
\input sec3
\input pbt
\input elab

@@

\bibliography{l}
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
